Updated Criteria for the Diagnostic Procedure for Parkinson’s Disease Dementia on Level I and their Validity in Deep Brain Stimulation Cohort

### 2025-05-07 _manuscript/manuscript.qmd
Moved algorithm-derivation and estimands to the Appendix.

library(targets)
library(tidyverse)

upstore <- here::here(tar_config_get("store"))
tar_source(here::here("R"))
data <- tar_read(raw_data, store = upstore)
crit <- tar_read(pdd_data, store = upstore)$criteria

@@

Suggestion: (MK) **(RO1) To estimate the prevalence of Parkinson’s disease dementia (PDD) and evaluate the diagnostic variability and concordance across different PDD criteria. (RO2) To identify specific diagnostic components that contribute to variability in PDD classification across the applied criteria.**^[**JM:Nice! Let's state it this way.**]

@@

## Theoretical and Empirical Estimands^[**JM: Put most of it to Appendix.**]

*Following the framework proposed by Lundberg et al. (2021), this section connects our research objectives and their corresponding theoretical (i.e., targets of inference) and empirical (i.e., data-driven) estimands. The theoretical estimand refers to a unit-specific quantity defined over a target population and represents the ideal quantity that would address the research question under optimal conditions, such as access to complete population data or perfect experimental control. In contrast, the empirical estimand corresponds to the quantity that is actually computable using the available dataset, given real world constraints. In our study, the target population for all theoretical estimands is defined as individuals diagnosed with Parkinson’s disease (PD) who are potential candidates for deep brain stimulation (DBS) treatment.*

*The theoretical estimand corresponding to the RO1 is the true prevalence of probable PDD in this population. Empirically, this is estimated by calculating the proportion of patients classified as having probable PDD according to each of the diagnostic criteria under consideration.*

*For the RO2, the theoretical estimand is the variance in prevalence attributable to the diagnostic process itself—that is, the extent to which different sets of criteria yield differing prevalence estimates for the same population. This variability is empirically quantified by comparing the distribution of PDD classifications across criteria within the cohort.*

*The RO3 concerns diagnostic concordance. Here, the theoretical estimand is a set of population-level contingency tables comparing PDD classifications assigned by each pair of diagnostic criteria, along with derived metrics such as sensitivity, specificity, and kappa coefficients. The corresponding empirical estimands are represented by matrices generated through pairwise Receiver Operating Characteristic (ROC) curve analyses that evaluate the discriminatory performance of each diagnostic set.*

*Finally, for the RO4, the theoretical estimand is defined as the set of diagnostic components whose variation systematically alters the probability of a probable PDD diagnosis. This aspect of the study is exploratory in nature. Empirically, we assess the contribution of each diagnostic feature by examining how variations in operational definitions (e.g., domain-specific thresholds, criteria for functional impairment) influence the empirical estimands derived for the first three objectives. This allows us to identify the diagnostic elements most responsible for inter-criterion discrepancies. We approach this RO from an exploratory point of view and evaluate the importance of each PDD criteria component by observing change in empirical estimands for RO1-3 when stratified by different oprationalization decisions.*

## Operationalization of Parkinson's Disease Dementia^[**JM: Add a section with some logic (Cartezian product) to Appendix as well.**]

In this study, we applied three distinct sets of diagnostic criteria for **probable**[^1] PDD at Level I. The first set was based on the original framework [@dubois2007], which utilized the Mini-Mental State Examination (MMSE) as a global cognitive screening tool, supplemented by assessments of attention, executive function, visuospatial abilities, and memory. The second set of criteria was drawn from the recent call for change of dementia diagnostic guidelines [@kulisevsky2024], which advocate for more sensitive cognitive domain assessments in the context of Parkinson's disease (PD). This updated approach incorporated specific items from the Montreal Cognitive Assessment (MoCA) to better detect PD-related dementia. The third approach applied the Czech version of the shortened Montreal Cognitive Assessment (sMoCA) [@bezdicek2020], a time-efficient modification designed to ascertain whether equivalent cognitive impairments could be reliably identified using a reduced testing protocol. Lastly, the fourth approach followed the Level II battery protocol, which is commonly used in the evaluation of PD-MCI. **The Level II methodology, including the use of a regression-based normative scoring approach, has been detailed in a prior study [@bezdicek2017]. Refer to @tbl-criteria**[^2] to a summary of the components and scoring thresholds of each diagnostic criterion. All non-cognitive criteria of probable PDD (i.e., diagnosis of PD that developed before dementia and absence of Major Depression, delirium or other abnormalities that obscure diagnosis were established by an independent neurological and psychiatric assessments and held true for all patients in the sample).

[^1]: **JM: We should unite the terminlogy. Most importantly, I reckon, we should differentiate between "probable PDD" (what is implied by meeting all the criteria within an operationalization) and "PDD" (the latent state of patient's cognition); and to use consistently the following trifecta - "criterion," "operationalization" and their "component".**

[^3]: **JM: We should unite the terminology. Most importantly, I reckon, we should differentiate between "probable PDD" (what is implied by meeting all the criteria within an operationalization) and "PDD" (the latent state of patient's cognition); and to use consistently the following trifecta - "criterion," "operationalization" and their "component".**

[^2]: **JM: Placeholder until we make the table proper.**

```{r}
#| label: "tbl-criteria"

gt::gt(crit)
```

@@

For each of these diagnostic approaches, we applied two operationalization strategies based on deficits in Instrumental Activities of Daily Living (IADL). First, we utilized FAQ item 9, which approximates the pill questionnaire from the original criteria [@dubois2007] employing a cut-off score of 2 points or higher. Second, we applied the entire Functional Activities Questionnaire (FAQ) as suggested in the call for change [@kulisevsky2024], employing a cut-off score of 7 points or higher based on Czech normative data [@bezdicek2011]. These methodologies resulted in a total of `r sum(critnum)` operationalizations, which were distributed across different diagnostic criteria: `r critnum["mmse"]` MMSE-based, `r critnum["moca"]` MoCA-based, `r critnum["smoca"]` sMoCA-based, and `r critnum["lvlII"]` based on the Level II battery.

@@

For sample description, we summarized continuous variables using mean, standard deviation, median, minimum, and maximum values. Categorical variables were summarized by the number of patients in each category. To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD operationalization listed in @tbl-criteria resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (operationalizations) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an operationalization (column). PDD prevalence estimates were then computed as $\frac{N_{PDD}}{N_{total}}$ separately for each operationalization to address RO1.

To address RO2-4, a set of two class cross-tabulations with associated statistics was computed for each pair of operationalizations via the `confusionMatrix()` function from the R package *caret* [@kuhn2008]. For each pair of operationalizations, the analysis was repeated twice such that each variables of the pair served once as the reference and once as the predictor. Following measures were used to evaluate pairwise concordance between different operationalizations of PDD criteria: 1) Cohen's $\kappa$ with its 95% confidence interval (CI) computed via the `cohen.kappa()` function from the R package *psych* [@revelle2024]; 2) Accuracy (i.e., the proportion of correct predictions, both true positives and true negatives, among the total number of cases) with its 95% CI; 3) Sensitivity/Recall (i.e., the proportion of true positives); and 4) Specificity (i.e., the proportion of true negatives).[^4]

@@

desc <- tar_read(sample_description    , store = upstore)
prev <- tar_read(prevalence_summaries  , store = upstore)
conc <- tar_read(concordance_statistics, store = upstore)
perc <- prev$table$perc
```

@@

```{r}
#| label: "tbl-desc"

@@

## Prevalence Estimates

Operationalization-wise prevalence estimates are presented in @tbl-prev. On average, estimated prevalence was `r do_summary(perc, 2, "M")`% (SD = `r do_summary(perc, 2, "SD")`, Md = `r do_summary(perc, 2, "Md")`, range `r do_summary(perc, 2, "minmax")`). Notably, the prevalence estimate was substantially lower when FAQ item 9 was used as a criterion of IADL deficit (M = `r do_summary(perc[grepl("FAQ 9", prev$table$IADL)], 2, "M")`%, SD = `r do_summary(perc[grepl("FAQ 9", prev$table$IADL)], 2, "SD")`, range `r do_summary(perc[grepl("FAQ 9", prev$table$IADL)], 2, "minmax")`) compared to using total FAQ score criterion (M = `r do_summary(perc[!grepl("FAQ 9", prev$table$IADL)], 2, "M")`%, SD = `r do_summary(perc[!grepl("FAQ 9", prev$table$IADL)], 2, "SD")`, range `r do_summary(perc[!grepl("FAQ 9", prev$table$IADL)], 2, "minmax")`) as demonstrated also in @fig-prev.[^7]

@@

\[Insert Table 3 and Figure 1[^8] here\]

[^8]: **JM: For thyself - add description.**

```{r}
#| label: "tbl-prev"

prev$gtable
```

```{r}
#| label: "fig-prev"

@@

>>>>

library(targets)
library(tidyverse)
library(gt)

upstore <- here::here(tar_config_get("store"))
tar_source(here::here("R"))
data <- tar_read(raw_data, store = upstore)
crit <- tar_read(pdd_data, store = upstore)$criteria
vars <- tar_read(variables, store = upstore)

@@

Suggestion: (MK) **(RO1) To estimate the prevalence of Parkinson’s disease dementia (PDD) and evaluate the diagnostic variability and concordance across different operationalizations of probable PDD. (RO2) To identify specific diagnostic components that contribute to variability in PDD classification across the applied criteria.**^[**JM:Nice! Let's state it this way.**]

@@

## Diagnostic algorithms for probable Parkinson's Disease Dementia^[**JM: Add a section with some logic (Cartezian product) to Appendix as well.**]

In this study, we applied three distinct sets of diagnostic criteria for probable PDD at Level I. The first set was based on the original framework [@dubois2007], which utilized the Mini-Mental State Examination (MMSE) as a global cognitive screening tool, supplemented by assessments of attention, executive function, visuospatial abilities, and memory. The second set of criteria was based on the recent call for change of dementia diagnostic guidelines [@kulisevsky2024], which advocate for more sensitive cognitive domain assessments in the context of PD. This updated approach incorporated specific items from the Montreal Cognitive Assessment (MoCA). The third approach applied the Czech version of the shortened Montreal Cognitive Assessment (sMoCA) [@bezdicek2020], a time-efficient modification designed to measure global cognitive performance using a reduced testing protocol that omits items providing redundant information. Lastly, the fourth approach followed the Level II protocol for diagnosis of PDD and Mild Cognitive Impairment in PD (PD-MCI) [@dubois2007; @litvan2012]. The Level II methodology, including the use of a regression-based normative scoring approach, has been detailed in a prior study [@bezdicek2017]. In this study, the threshold for cognitive impairment at Level II was set at $z \leq -1.5$. All non-cognitive criteria of probable PDD (i.e., diagnosis of PD that developed before dementia and absence of Major Depression, delirium or other abnormalities that obscure diagnosis were established by an independent neurological and psychiatric assessments) held true for all patients in the sample according to the psychiatric and neurological examinations.

```{r}
#| label: "tbl-criteria"
#| tbl-cap: "Summary of probable PDD operationalizations compared in the study."

table_criteria(vars)
```

@@

For each of these diagnostic approaches, we applied two operationalization strategies based on deficits in Instrumental Activities of Daily Living (IADL). First, we utilized FAQ item 9, which approximates the pill questionnaire from the original criteria [@dubois2007] employing a cut-off score of 2 points or higher. Second, we applied the entire Functional Activities Questionnaire (FAQ) as suggested in the call for change [@kulisevsky2024], employing a cut-off score of 7 points or higher based on Czech normative data [@bezdicek2011]. These methodologies resulted in a total of `r sum(critnum)` algorithms, which were distributed across different diagnostic criteria: `r critnum["mmse"]` MMSE-based, `r critnum["moca"]` MoCA-based, `r critnum["smoca"]` sMoCA-based, and `r critnum["lvlII"]` based on the Level II battery (see @tbl-criteria and Appendix for the exact specification of each algorithm).

@@

Following the framework proposed by Lundberg et al. (2021), in this study we explicitly connect our research objectives and their corresponding theoretical (i.e., targets of inference) and empirical (i.e., data-driven) estimands to statistical estimates. The theoretical estimand refers to a unit-specific quantity defined over a target population and represents the ideal quantity that would address the research question under optimal conditions, such as access to complete population data or perfect experimental control. In contrast, the empirical estimand corresponds to the quantity that is actually computable using the available dataset, given real world constraints. The full description of the study's estimands and their relation to our research objectives is presented in the Appendix (see @tbl-estimands).

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD operationalization listed in @tbl-criteria resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (operationalizations) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an operationalization (column). PDD prevalence estimates were then computed as $\frac{N_{PDD}}{N_{total}}$ separately for each operationalization to address RO1. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each operationalization of probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

Next, a set of two class cross-tabulations with associated statistics was computed for each pair of operationalizations via the `confusionMatrix()` function from the R package *caret* [@kuhn2008]. For each pair of operationalizations, the analysis was repeated twice such that each variables of the pair served once as the reference and once as the predictor. Following measures were used to evaluate pairwise concordance between different operationalizations of PDD criteria: 1) Cohen's $\kappa$ with its 95% confidence interval (CI) computed via the `cohen.kappa()` function from the R package *psych* [@revelle2024]; 2) Accuracy (i.e., the proportion of correct predictions, both true positives and true negatives, among the total number of cases) with its 95% CI; 3) Sensitivity/Recall (i.e., the proportion of true positives); and 4) Specificity (i.e., the proportion of true negatives).[^4]

@@

desc <- tar_read(sample_description, store = upstore)
prev <- tar_read(prevalence_summaries, store = upstore)
conc <- tar_read(concordance_statistics, store = upstore)
perc <- prev$table$perc
```

@@

```{r}
#| label: "tbl-desc"
#| tbl-cap: "Demographic, clinical and cognitive characteristics of the sample."

@@

## PDD Rate Estimates

Operationalization-wise rate of PDD estimates are presented in @tbl-rates. On average, estimated PDD rate was `r do_summary(perc, 2, "M")`% (SD = `r do_summary(perc, 2, "SD")`, Md = `r do_summary(perc, 2, "Md")`, range `r do_summary(perc, 2, "minmax")`). Notably, the estimate was substantially lower when FAQ item 9 was used as a criterion of IADL deficit (M = `r do_summary(perc[grepl("FAQ 9", prev$table$IADL)], 2, "M")`%, SD = `r do_summary(perc[grepl("FAQ 9", prev$table$IADL)], 2, "SD")`, range `r do_summary(perc[grepl("FAQ 9", prev$table$IADL)], 2, "minmax")`) compared to using total FAQ score criterion (M = `r do_summary(perc[!grepl("FAQ 9", prev$table$IADL)], 2, "M")`%, SD = `r do_summary(perc[!grepl("FAQ 9", prev$table$IADL)], 2, "SD")`, range `r do_summary(perc[!grepl("FAQ 9", prev$table$IADL)], 2, "minmax")`) as demonstrated also in @fig-prev.[^7]

```{r}
#| label: "tbl-rates"
#| tbl-cap: "Estimates of the rate of probable PDD in the sample."

prev$gtables$rates
```

@@

\[Insert Table 3 & Figure 1 here\]

```{r}
#| label: "fig-prev"
#| fig-cap: "Summary of the estimates of probable PDD rate across operationalizations under consideration. Vertical lines represent estimates arrived at by using sMoCA (dotted) or Level II (dashed) with FAQ item 9 (orange) or FAQ total score (blue) as criteria for probable PDD."

@@

## Derivation of the Algorithms Set

Both, the original PDD criteria [@dubois2007] and the call for their change [@kulisevsky2024] allow for several distinct combinations of test indexes to be used to define cognitive impairment. Consequently, in this study we derived all algorithms for probable PDD on Level I that are in line with published criteria. This procedure parallel the diagnostic algorithm outlined in Table 2 of @dubois2007. Specifically, in this study, we varied the exact specification of items 3-5 of this table (i.e., the measure of global cognitive impairment, the measure of the impact on ADLs and the measures of impaired cognition).

For each set of criteria (MMSE-based, MoCA-based, sMoCA-based and Level II) we first specified measures and then the thresholds for each measure used to define probable PDD. Wherever more than one option was present in either the choice of measure or the choice of threshold, we created an algorithm for each choice in turn. The final set of algorithms tested was arrived at by computing the Cartesian product of all possibilities provided by varying measures and thresholds. All combinations are presented in @tbl-algos.

For MMSE-based algorithms, the following sets of items served as the basis:

$Global = \{MMSE < 26\}$

$Attention = \{{Sevens\ backwards < 4}\}$

$Executive = \{{Clock\ drawing} < 2, {Lexical\ fluency\ (S)} < 10\}$

$Construction = \{Pentagons < 1\}$

$Memory = \{{Three{\text-}words\ recall} < 3\}$

$IADL = \{FAQ > 7, {FAQ\ (it. 9)} > 1\}$

The ensuing Cartesian product $Global \times Attention \times Executive \times Construction \times Memory \times IADL$ results in $1 \times 1 \times 2 \times 1 \times 1 \times 2 = 4$ MMSE-based algorithms for probable PDD.

For MoCA-based algorithms, the following sets of items served as the basis:

$Global = \{MoCA < 27\}$

$Attention = \{{Sevens\ backwards < 3}\}$

$Executive = \{{Clock\ drawing} < 2, {Clock\ drawing} < 3, {Lexical\ fluency\ (K)} < 11\}$

$Construction = \{Cube\ drawing < 1\}$

$Memory = \{Five{\text-}words\ recall < 1, Five{\text-}words\ recall < 2, Five{\text-}words\ recall < 3, Five{\text-}words\ recall < 4, Five{\text-}words\ recall < 5\}$

$Language = \{Abstraction < 2, Animal\ naming < 3\}$

$IADL = \{FAQ > 7, {FAQ\ (it. 9)} > 1\}$

Note that the additional language domain adds complexity to establishing a diagnostic algorithm because simply by adding it to the set, the number of potential algorithm doubles. Further complexity is added by the fact that there are so far no guidelines for selecting a diagnostic threshold for the Clock drawing and Five-words recall tests, both of which differ from their MMSE counterparts. Computing the Cartesian product $Global \times Attention \times Executive \times Construction \times Memory \times Language \times IADL$ yields $1 \times 1 \times 3 \times 1 \times 5 \times 2 \times 2 = 60$ distinct MoCA-based algorithms for probable PDD.

For sMoCA-based algorithms, the following sets of items served as the basis:

$Global = \{sMoCA < 13\}$

$IADL = \{FAQ > 7, {FAQ\ (it. 9)} > 1\}$

yielding $Global \times IADL$ yields $1 \times 2 = 2$ distinct sMoCA-based algorithms for probable PDD.

Finally, the Level II algorithms were based on the following sets of items:

$Attention = \{z(TMT\ A) < -1.5\ \cup z(WAIS\ DSB) < -1.5\}$

$Executive = \{z(CF\ A) < -1.5\ \cup z(PST\ C) < -1.5\}$

$Construction = \{z(JoLO) < -1.5\ \cup z(CLOXI) < -1.5\}$

$Memory = \{z(RAVLT\ DR) < -1.5\ \cup z(BVMTR\ DR) < -1.5\ \cup z(WMS\text-III\ Family\ Pictures) < -1.5\}$

$Language = \{z(WAIS\ Similarities) < -1.5 \ \cup z(BNT\ 60) < -1.5\}$

$IADL = \{FAQ > 7, {FAQ\ (it. 9)} > 1\}$

where $z()$ denotes calculation of age, sex and education adjusted z-score. This yields $1 \times 1 \times 1 \times 1 \times 1 \times 2 = 2$ distinct Level II algorithms for probable PDD in the current study.

```{r}
#| label: "tbl-algos"
#| tbl-cap: "Summary of all algortihms for probable PDD used in the study."

prev$gtables$algorithms
```

## Theoretical and Empirical Estimands

In this study, we follow the framework proposed by Lundberg et al. (2021) for specifying targets of inference (i.e., the estimands) in qunatitative science to increase transparency and connect our statistical evidence to relevant theory. @tbl-estimands contains verbal description of the components relating to each of our proclaimed research objectives and map them to the population quantity of interest (the theoretical estimand), data-dependent quantity that could be estimated (the empirical estimand) and quantities that are reported in the study (statistical estimates).

```{r}
#| label: "tbl-estimands"
#| tbl-cap: "Mapping between research objectives and quantities of interest in the current study."

table_estimands()
```

The RO1, to estimate the prevalence of PDD and evaluate the diagnostic variability and concordance across different operationalizations of probable PDD, was further subdivided into four distinct research objectives - to estimate the rate of PDD within PD (RO1.1), to estimated variability of this rate (RO1.2), to evaluate predictive value of demographic variables for probable PDD classification (RO1.3) and to evaluate concordance between different probable PDD operationalizations and criteria (RO1.4). Estimates relating to RO1.1 and RO1.3 cannot be safely generalized beyond a population of PD patients that are candidates for DBS due to the systematic differences between DBS candidates pool and general PD population (such as lower age of DBS candidates compared to the general PD population). On the other hand, the estimates relating to RO1.4 (and to a lesser degree to RO1.2) may not be substantially influenced by the sample at hand as the primary source of their variance comes from variability in measures employed (e.g., MMSE vs MoCA to assess global cognitive performance) rather than variability in subjects' performance. Assuming that there is no substantial Differential Item Functioning between DBS candidates and broader PD population, the estimates relating to RO1.4 can be cautiously generalized beyond the current sample.

Finally, for the RO2, the theoretical estimand is defined as the set of diagnostic components whose variation systematically alters the probability of a probable PDD diagnosis. This aspect of the study is exploratory in nature. Empirically, we assess the contribution of each diagnostic feature by examining how variations in operational definitions (e.g., domain-specific thresholds, criteria for functional impairment) influence the statistical estimates derived for the first objective. This allows us to identify the diagnostic elements most responsible for inter-operationalization discrepancies.

## Supplementary Presentation of Results
